---
title: "HW 01"
author: "Sravani Maddineni"
format: 
  html:
    embed-resources: true
execute :
  warning: false
  message: false
  error: false
toc: true
---

## 0 - Setup

```{r setup}
if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse,
               palmerpenguins) 

devtools::install_github("tidyverse/dsbox")
```

```{r}

```

## 1 - Road traffic accidents in Edinburgh

```         
```

```{r}
# Load and glimpse the dataset
library(tidyverse)
library(hms)

accidents <- read_csv("data/accidents.csv")

# View structure: number of rows, columns, and variable types
glimpse(accidents)

```

```{r}
# Preprocessing: convert date/time and classify as weekend/weekday
accidents <- accidents %>%
  mutate(
    date = as.Date(date),
    time = as_hms(time),
    day_type = if_else(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"),
    severity = factor(severity, levels = c("Fatal", "Serious", "Slight"))
  )

```

```{r}
# Density plot showing accident times by day type and severity
ggplot(accidents, aes(x = time, fill = severity)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~day_type, ncol = 1) +
  scale_fill_manual(values = c("Fatal" = "purple", "Serious" = "steelblue", "Slight" = "khaki")) +
  labs(
    title = "Accident Density in Edinburgh Throughout the Day (2018)",
    subtitle = "By Day Type (Weekend vs Weekday) and Severity",
    x = "Time of Day",
    y = "Density",
    fill = "Severity"
  ) +
  theme_minimal()

```

Interpretation

The dataset contains records of traffic accidents in Edinburgh during 2018. After inspecting the structure with glimpse( ), we see variables such as date, time, location, and severity.

The plot above shows the distribution of accident times across weekdays and weekends. Most accidents happen during morning and evening hours, likely aligned with commuting times. Slight accidents dominate, but serious and fatal ones also peak around the same hours. Weekends show a delay in peak accident time, indicating different travel behavior.

```         
```

## 2 - NYC marathon winners

```         
```

## 3 - US counties

```{r}

# Incorrect combined plot
ggplot(USA) +
  geom_point(aes(x = median_edu, y = median_hh_income)) +
  geom_boxplot(aes(x = smoking_ban, y = pop2017))

# Facet row
ggplot(county %>% filter(!is.na(median_edu))) +
  geom_point(aes(x = homeownership, y = poverty)) +
  facet_grid(median_edu ~ .)

# Facet column
ggplot(county %>% filter(!is.na(median_edu))) +
  geom_point(aes(x = homeownership, y = poverty)) +
  facet_grid(. ~ median_edu)

# Plot A to H
plots <- list(
  ggplot(county, aes(x = homeownership, y = poverty)) + geom_point(),
  ggplot(county, aes(x = homeownership, y = poverty)) + geom_point() + geom_smooth(color = "blue", se = FALSE),
  ggplot(county, aes(x = homeownership, y = poverty)) + geom_point() +
    geom_smooth(aes(group = metro), method = "loess", span = 1.4, color = "green", se = FALSE),
  ggplot(county, aes(x = homeownership, y = poverty)) +
    geom_smooth(aes(group = metro), method = "loess", span = 1.4, color = "blue", se = FALSE) + geom_point(),
  ggplot(county, aes(x = homeownership, y = poverty, color = metro)) +
    geom_point() + geom_smooth(aes(group = metro, linetype = metro), method = "loess", span = 1.4, color = "blue", se = FALSE),
  ggplot(county, aes(x = homeownership, y = poverty, color = metro)) +
    geom_point() + geom_smooth(aes(group = metro, color = metro), method = "loess", span = 1.4, se = FALSE),
  ggplot(county, aes(x = homeownership, y = poverty, color = metro)) +
    geom_point() + geom_smooth(color = "blue", se = FALSE),
  ggplot(county, aes(x = homeownership, y = poverty, color = metro)) + geom_point()
)

```

## 4 - Rental apartments in SF

## 5 - Napoleon’s march.
